---
title: "A Replication of Karlan and List (2007)"
author: "Rishikumar Mathiazhagan"
date: today
callout-appearance: minimal # this hides the blue "i" icon on .callout-notes
---


## Introduction

Dean Karlan at Yale and John List at the University of Chicago conducted a field experiment to test the effectiveness of different fundraising letters. They sent out 50,000 fundraising letters to potential donors, randomly assigning each letter to one of three treatments: a standard letter, a matching grant letter, or a challenge grant letter. They published the results of this experiment in the _American Economic Review_ in 2007. The article and supporting data are available from the [AEA website](https://www.aeaweb.org/articles?id=10.1257/aer.97.5.1774) and from Innovations for Poverty Action as part of [Harvard's Dataverse](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/27853&version=4.2).

To better understand the demand side of charitable giving, Karlan and List used a natural field experiment involving over 50,000 previous donors to a politically liberal nonprofit organization. The letters were randomized to test three main dimensions: whether a matching grant was offered, the ratio of the match (1:1, 2:1, or 3:1), and the size of the matching fund ($25,000, $50,000, $100,000, or unstated). In addition, suggested donation amounts were varied using multipliers of the recipient’s previous highest contribution. Their findings revealed that simply offering a match significantly increased both the likelihood of donating and the average amount raised per letter. However, increasing the match ratio beyond 1:1 did not yield additional benefits, challenging conventional fundraising wisdom. Notably, the effect of matching offers was more pronounced in conservative-leaning ("red") states, suggesting that political context may shape donor responsiveness.

This project seeks to replicate their results.


## Data

### Description

```{r}
library(haven)
library(dplyr)
library(summarytools)

data <- read_dta("karlan_list_2007.dta")

glimpse(data)

dfSummary(data)
```

:::: {.callout-note collapse="true"}
### Variable Definitions

| Variable             | Description                                                         |
|----------------------|---------------------------------------------------------------------|
| `treatment`          | Treatment                                                           |
| `control`            | Control                                                             |
| `ratio`              | Match ratio                                                         |
| `ratio2`             | 2:1 match ratio                                                     |
| `ratio3`             | 3:1 match ratio                                                     |
| `size`               | Match threshold                                                     |
| `size25`             | \$25,000 match threshold                                            |
| `size50`             | \$50,000 match threshold                                            |
| `size100`            | \$100,000 match threshold                                           |
| `sizeno`             | Unstated match threshold                                            |
| `ask`                | Suggested donation amount                                           |
| `askd1`              | Suggested donation was highest previous contribution                |
| `askd2`              | Suggested donation was 1.25 x highest previous contribution         |
| `askd3`              | Suggested donation was 1.50 x highest previous contribution         |
| `ask1`               | Highest previous contribution (for suggestion)                      |
| `ask2`               | 1.25 x highest previous contribution (for suggestion)               |
| `ask3`               | 1.50 x highest previous contribution (for suggestion)               |
| `amount`             | Dollars given                                                       |
| `gave`               | Gave anything                                                       |
| `amountchange`       | Change in amount given                                              |
| `hpa`                | Highest previous contribution                                       |
| `ltmedmra`           | Small prior donor: last gift was less than median \$35              |
| `freq`               | Number of prior donations                                           |
| `years`              | Number of years since initial donation                              |
| `year5`              | At least 5 years since initial donation                             |
| `mrm2`               | Number of months since last donation                                |
| `dormant`            | Already donated in 2005                                             |
| `female`             | Female                                                              |
| `couple`             | Couple                                                              |
| `state50one`         | State tag: 1 for one observation of each of 50 states; 0 otherwise  |
| `nonlit`             | Nonlitigation                                                       |
| `cases`              | Court cases from state in 2004-5 in which organization was involved |
| `statecnt`           | Percent of sample from state                                        |
| `stateresponse`      | Proportion of sample from the state who gave                        |
| `stateresponset`     | Proportion of treated sample from the state who gave                |
| `stateresponsec`     | Proportion of control sample from the state who gave                |
| `stateresponsetminc` | stateresponset - stateresponsec                                     |
| `perbush`            | State vote share for Bush                                           |
| `close25`            | State vote share for Bush between 47.5% and 52.5%                   |
| `red0`               | Red state                                                           |
| `blue0`              | Blue state                                                          |
| `redcty`             | Red county                                                          |
| `bluecty`            | Blue county                                                         |
| `pwhite`             | Proportion white within zip code                                    |
| `pblack`             | Proportion black within zip code                                    |
| `page18_39`          | Proportion age 18-39 within zip code                                |
| `ave_hh_sz`          | Average household size within zip code                              |
| `median_hhincome`    | Median household income within zip code                             |
| `powner`             | Proportion house owner within zip code                              |
| `psch_atlstba`       | Proportion who finished college within zip code                     |
| `pop_propurban`      | Proportion of population urban within zip code                      |

::::


### Balance Test 

As an ad hoc test of the randomization mechanism, I provide a series of tests that compare aspects of the treatment and control groups to assess whether they are statistically significantly different from one another.

```{r}

library(haven)
library(dplyr)
library(broom)

# Load dataset
data <- read_dta("karlan_list_2007.dta")

# Split data by treatment
treat <- filter(data, treatment == 1)
control <- filter(data, treatment == 0)

# Variables to test
vars <- c("mrm2", "hpa", "freq","page18_39","red0")

# Function to report results
for (var in vars) {
  cat("\n==============================\n")
  cat("Variable:", var, "\n")
  
  # Means and SDs
  treat_mean <- mean(treat[[var]], na.rm = TRUE)
  control_mean <- mean(control[[var]], na.rm = TRUE)
  treat_sd <- sd(treat[[var]], na.rm = TRUE)
  control_sd <- sd(control[[var]], na.rm = TRUE)
  overall_mean <- mean(data[[var]], na.rm = TRUE)
  overall_sd <- sd(data[[var]], na.rm = TRUE)
  
  cat("Treatment Mean:", round(treat_mean, 3), "SD:", round(treat_sd, 3), "\n")
  cat("Control Mean:  ", round(control_mean, 3), "SD:", round(control_sd, 3), "\n")
  
  # T-test
  ttest <- t.test(as.formula(paste(var, "~ treatment")), data = data)
  cat("T-test p-value:", round(ttest$p.value, 4), "\n")
  
  # Regression
  model <- lm(as.formula(paste(var, "~ treatment")), data = data)
  coef_summary <- tidy(model) %>% filter(term == "treatment")
  cat("Regression Coefficient:", round(coef_summary$estimate, 4), "\n")
  cat("Regression p-value:    ", round(coef_summary$p.value, 4), "\n")
}
```
All tested variables show very similar means and standard deviations across treatment and control groups, as expected under successful randomization. The t-test and regression p-values are nearly identical and all above 0.05, indicating no statistically significant imbalance. These results replicate Table 1 in the Karlan & List (2007) paper, which is included to show that randomization produced comparable groups on observable characteristics.


## Experimental Results

### Charitable Contribution Made

First, I analyze whether matched donations lead to an increased response rate of making a donation. 

```{r}
library(haven)
library(dplyr)
library(ggplot2)

# Load data
data <- read_dta("karlan_list_2007.dta")

# Calculate proportion who donated by treatment group
response_rate <- data %>%
  group_by(treatment) %>%
  summarise(proportion_donated = mean(gave, na.rm = TRUE)) %>%
  mutate(group = ifelse(treatment == 1, "Treatment", "Control"))

# Plot
ggplot(response_rate, aes(x = group, y = proportion_donated, fill = group)) +
  geom_col(width = 0.5) +
  labs(
    title = "Proportion of People Who Donated",
    x = "",
    y = "Proportion",
    fill = ""
  ) +
  ylim(0, max(response_rate$proportion_donated) * 1.1) +
  theme_minimal()
```

```{r}

library(haven)
library(dplyr)
library(broom)

# Load data
data <- read_dta("karlan_list_2007.dta")

# T-test
ttest_gave <- t.test(gave ~ treatment, data = data)
cat("T-test result:\n")
print(ttest_gave)

# Regression
model_gave <- lm(gave ~ treatment, data = data)
reg_summary <- tidy(model_gave)
cat("\nLinear regression result:\n")
print(reg_summary)
```
The t-test comparing the proportion of individuals who made any charitable donation between the treatment and control groups shows a small but statistically significant increase in response rate in the treatment group. This confirms the finding in Table 2A, Panel A, where the response rate increases from 1.8% to 2.2% when matching is introduced.

The bivariate linear regression yields an equivalent result: the coefficient on `treatment` is positive and significant, indicating that simply offering a match increases the likelihood of donation.

**Interpretation:** Even a small increase in donation probability — from just under 2% to just over 2% — is meaningful in a large fundraising campaign. This suggests that people are psychologically responsive to matching offers, even when the absolute value of the match is not very large. The perception of increased impact ("my $10 becomes $20") may be enough to nudge more people into giving. This supports the idea that **framing and context matter in charitable behavior**, not just individual preferences.

```{r}

library(haven)
library(dplyr)
library(broom)
library(margins)

# Load data
data <- read_dta("karlan_list_2007.dta")

# Probit regression: gave ~ treatment
probit_model <- glm(gave ~ treatment, data = data, family = binomial(link = "probit"))

# Summary of the model
summary(probit_model)

# Tidy output (for coefficient and SE)
tidy(probit_model)

# Optional: Marginal effects (not required for Table 3, but helpful)
margins::margins(probit_model)

model_full <- glm(
  gave ~ treatment * ratio2 +
         treatment * ratio3 +
         treatment * size25 +
         treatment * size50 +
         treatment * size100 +
         treatment * askd2 +
         treatment * askd3,
  data = data,
  family = binomial(link = "probit")
)

library(broom)
tidy(model_full)
```

### Interpretation of Probit Regression Result

The probit regression produced a positive and statistically significant coefficient on the treatment variable, with an estimated effect of **0.0868**, a z-value of **3.11**, and a p-value below **0.01**. This confirms that assignment to the treatment group — those who received a matching donation offer — was associated with a higher probability of making a donation.

The result is statistically significant at the **1% level**, providing strong evidence that the offer of a match influenced donor behavior. While the numerical value of the coefficient may seem modest, its significance within a nonlinear probit framework reinforces the idea that behavioral prompts like matching gifts can effectively shape outcomes.

In essence, this supports the broader finding from Karlan and List (2007): even relatively small psychological nudges can meaningfully impact decision-making in a real-world charitable context.


### Differences between Match Rates

Next, I assess the effectiveness of different sizes of matched donations on the response rate.


```{r}

library(haven)
library(dplyr)

# Load data
data <- read_dta("karlan_list_2007.dta")

# Filter only the people who were in a match group (i.e., treatment == 1)
matched <- filter(data, treatment == 1)

# Create a variable for 1:1 match (implied if ratio2 and ratio3 are both 0)
matched <- matched %>%
  mutate(ratio1 = ifelse(ratio2 == 0 & ratio3 == 0, 1, 0))

# T-test: 2:1 match vs 1:1 match
t_21_vs_11 <- t.test(gave ~ ratio2, data = filter(matched, ratio1 == 1 | ratio2 == 1))

# T-test: 3:1 match vs 1:1 match
t_31_vs_11 <- t.test(gave ~ ratio3, data = filter(matched, ratio1 == 1 | ratio3 == 1))

# Output results
cat("T-test: 2:1 vs 1:1 match rate\n")
print(t_21_vs_11)

cat("\nT-test: 3:1 vs 1:1 match rate\n")
print(t_31_vs_11)
```

### Effectiveness of Different Match Ratios: T-Test Results

To evaluate whether more generous match ratios (2:1 and 3:1) increase the likelihood of donating compared to the standard 1:1 match, I conducted two t-tests:

- **2:1 vs 1:1 match rate**
- **3:1 vs 1:1 match rate**

In both cases, the t-tests found **no statistically significant difference** in the probability of donation. The p-value for the 2:1 vs 1:1 comparison was **0.3345**, and for the 3:1 vs 1:1 comparison it was **0.3101** — both well above the conventional significance threshold of 0.05.

The estimated differences in means were very small, with confidence intervals that include zero, indicating no reliable effect of increasing the match ratio.

These results support the authors’ comment on page 8 of the paper that "**larger match ratios had no additional impact**." In other words, donors respond to the **presence** of a match offer, but increasing the match from 1:1 to 2:1 or 3:1 does not lead to a higher response rate.

This finding reinforces a key insight from behavioral economics: **perceived opportunity or framing (i.e., the existence of a match), rather than the size of the match, drives behavior** in charitable giving contexts.

```{r}

library(haven)
library(dplyr)
library(broom)

# Load data
data <- read_dta("karlan_list_2007.dta")

# Subset only treatment group (those who received any match)
matched <- filter(data, treatment == 1)

# Create ratio1 manually: it's the group not in ratio2 or ratio3
matched <- matched %>%
  mutate(ratio1 = ifelse(ratio2 == 0 & ratio3 == 0, 1, 0))

# Run regression with dummy variables (ratio1 is implicit baseline)
model_ratios <- lm(gave ~ ratio2 + ratio3, data = matched)

# Output tidy summary
tidy(model_ratios)
```

### Regression Analysis: Impact of Match Ratio on Donation Behavior

To assess whether the generosity of the match offer affects the likelihood of donating, I ran a linear regression using data from individuals who were in the treatment group. The model includes dummy variables for the 2:1 and 3:1 match offers, with the 1:1 match offer serving as the reference category.

The regression results show that:
- The coefficient on `ratio2` is **0.00188** with a standard error of **0.00197**, and a p-value of **0.338**.
- The coefficient on `ratio3` is **0.00198** with a standard error of **0.00197**, and a p-value of **0.313**.

These coefficients represent the difference in donation probability relative to the 1:1 match group. However, both coefficients are small and **statistically insignificant**, indicating that neither the 2:1 nor 3:1 match rate led to a meaningful increase in the likelihood of donating compared to a 1:1 match.

These findings are entirely consistent with earlier t-test results and support the authors’ conclusion that **increasing the match ratio has no additional effect**. The behavioral impact appears to stem from the presence of a match offer itself, rather than its magnitude. Donors may not be especially sensitive to whether the match is 1:1 or 3:1 — the idea of leverage seems to matter more than the amount.

```{r}

library(haven)
library(dplyr)
library(broom)

# Load data
data <- read_dta("karlan_list_2007.dta")

# Keep only treatment group
matched <- filter(data, treatment == 1)

# Create 1:1 dummy
matched <- matched %>%
  mutate(ratio1 = ifelse(ratio2 == 0 & ratio3 == 0, 1, 0))

# ----- APPROACH 1: Direct from the data -----

# Compute response rates by match group
response_rates <- matched %>%
  mutate(ratio_type = case_when(
    ratio1 == 1 ~ "1:1",
    ratio2 == 1 ~ "2:1",
    ratio3 == 1 ~ "3:1"
  )) %>%
  group_by(ratio_type) %>%
  summarise(response_rate = mean(gave, na.rm = TRUE))

# Extract values for comparison
rate_1_1 <- response_rates$response_rate[response_rates$ratio_type == "1:1"]
rate_2_1 <- response_rates$response_rate[response_rates$ratio_type == "2:1"]
rate_3_1 <- response_rates$response_rate[response_rates$ratio_type == "3:1"]

# Calculate differences
diff_21_11 <- rate_2_1 - rate_1_1
diff_31_21 <- rate_3_1 - rate_2_1

# ----- APPROACH 2: From regression coefficients -----

# Run regression using 1:1 (ratio1) as the baseline
model <- lm(gave ~ ratio2 + ratio3, data = matched)
model_out <- tidy(model)

# Coefficient difference (they are relative to 1:1 group)
coef_21_11 <- model_out$estimate[model_out$term == "ratio2"]
coef_31_21 <- model_out$estimate[model_out$term == "ratio3"] - coef_21_11

# Output results
cat("== Response Rate Differences from Raw Data ==\n")
cat("2:1 - 1:1 =", round(diff_21_11, 5), "\n")
cat("3:1 - 2:1 =", round(diff_31_21, 5), "\n")

cat("\n== Differences in Fitted Coefficients ==\n")
cat("2:1 vs 1:1 =", round(coef_21_11, 5), "\n")
cat("3:1 vs 2:1 =", round(coef_31_21, 5), "\n")
```

### Match Ratio Differences: Response Rate and Regression Analysis

To understand whether higher match ratios increase the likelihood of donation, I calculated response rate differences in two ways:

#### 1. Differences in Raw Response Rates
- **2:1 vs 1:1** match ratio: 0.00188
- **3:1 vs 2:1** match ratio: 0.00010

#### 2. Differences in Fitted Coefficients (Linear Model)
- **2:1 vs 1:1**: 0.00188
- **3:1 vs 2:1**: 0.00010

Both approaches yielded the **same differences**, confirming that the estimated impact of increasing the match ratio is extremely small — around **0.1 to 0.2 percentage points**.

### Conclusion
These findings provide strong evidence that increasing the match ratio from 1:1 to 2:1 or from 2:1 to 3:1 does **not significantly increase** the probability that someone donates. 

This aligns with the authors' conclusion: the **presence of a match offer**, rather than its **generosity**, is what motivates donations. Donors respond to the idea of a match, but **larger matches do not provide additional behavioral lift**.

In sum, the marginal effectiveness of more generous matches is negligible — a key insight for fundraisers seeking efficient campaign strategies.


### Size of Charitable Contribution

In this subsection, I analyze the effect of the size of matched donation on the size of the charitable contribution.

```{r}
library(haven)
library(dplyr)
library(broom)

# Load data
data <- read_dta("karlan_list_2007.dta")

# T-test: Compare average donation amount
t_test_amount <- t.test(amount ~ treatment, data = data)
cat("== T-test: Donation Amount by Treatment Status ==\n")
print(t_test_amount)

# Regression: Amount ~ Treatment
model_amount <- lm(amount ~ treatment, data = data)
cat("\n== Linear Regression: Amount ~ Treatment ==\n")
tidy(model_amount)
```

### Effect of Match Offers on Donation Amounts

To assess whether being offered a matching donation opportunity affects the amount individuals donate, I conducted both a t-test and a bivariate linear regression using the full dataset.

The **t-test** comparing average donation amounts between treatment and control groups yielded a **p-value of 0.055**, which is just above the traditional 5% significance threshold. The average donation was $0.81 in the control group and $0.97 in the treatment group — a difference of approximately $0.15.

The **linear regression** of donation amount on treatment status produced a coefficient of **0.154** with a **p-value of 0.0628**, again marginally above the 5% threshold. This suggests a small, positive effect of treatment on donation amount, but not strong enough to be deemed statistically significant at conventional levels.

### Interpretation

These results indicate that **being offered a match may slightly increase the average amount donated**, but the evidence is not strong enough to make a confident causal claim. The p-values suggest that the observed difference might be due to chance, though the direction of the effect (positive) is consistent with the broader narrative of the study.

In summary, while matching offers clearly increase the likelihood that someone donates, the evidence that they increase the **size** of the donation (unconditionally) is **suggestive but not definitive**.

```{r}

library(haven)
library(dplyr)
library(broom)

# Load the data
data <- read_dta("karlan_list_2007.dta")

# Filter to donors only (those who gave something)
donors_only <- filter(data, gave == 1)

# T-test: Compare average amount given among donors
t_test_donors <- t.test(amount ~ treatment, data = donors_only)
cat("== T-test: Donation Amount by Treatment Status (Donors Only) ==\n")
print(t_test_donors)

# Regression: Amount ~ Treatment (among donors)
model_donors <- lm(amount ~ treatment, data = donors_only)
cat("\n== Linear Regression: Amount ~ Treatment (Donors Only) ==\n")
tidy(model_donors)
```

### Conditional Effect of Treatment on Donation Amount (Among Donors Only)

To examine whether match offers influenced how much people donated — conditional on having donated — I restricted the dataset to only those individuals who gave a positive amount.

#### T-Test Results
The average donation among the **control group** was **$45.54**, while the average in the **treatment group** was **$43.87**. The t-test revealed **no statistically significant difference** between the two groups (p = 0.559). The 95% confidence interval for the difference in means includes zero, reinforcing this conclusion.

#### Regression Results
The linear regression of donation amount on treatment (among donors only) produced a coefficient of **-1.67**, indicating that those in the treatment group donated slightly less on average. However, the effect is **not statistically significant** (p = 0.561).

#### Interpretation and Causal Insights
These results suggest that **being offered a match does not influence the size of the donation**, once someone has decided to donate. In fact, if anything, treated individuals gave slightly **less**, though the difference is not meaningful.

Importantly, **this analysis is conditional on donation behavior**, which was itself affected by the treatment. Because we are conditioning on a **post-treatment variable** (gave), the regression coefficient **does not have a causal interpretation**. The group of donors in the treatment group may differ in unobservable ways from donors in the control group, and this could bias the estimate.

In summary, this analysis tells us that the **treatment affects the extensive margin (whether people give), not the intensive margin (how much they give)** — a conclusion that aligns with the broader findings of Karlan and List (2007).

```{r}

library(haven)
library(dplyr)
library(ggplot2)

# Load data
data <- read_dta("karlan_list_2007.dta")

# Filter to donors only
donors <- filter(data, gave == 1)

# Calculate group means
group_means <- donors %>%
  group_by(treatment) %>%
  summarise(mean_amount = mean(amount, na.rm = TRUE))

# Plot for Treatment Group
ggplot(filter(donors, treatment == 1), aes(x = amount)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "white") +
  geom_vline(xintercept = group_means$mean_amount[group_means$treatment == 1],
             color = "red", size = 1) +
  labs(
    title = "Donation Amounts (Treatment Group)",
    x = "Amount Donated",
    y = "Count"
  ) +
  theme_minimal()

# Plot for Control Group
ggplot(filter(donors, treatment == 0), aes(x = amount)) +
  geom_histogram(binwidth = 5, fill = "lightgreen", color = "white") +
  geom_vline(xintercept = group_means$mean_amount[group_means$treatment == 0],
             color = "red", size = 1) +
  labs(
    title = "Donation Amounts (Control Group)",
    x = "Amount Donated",
    y = "Count"
  ) +
  theme_minimal()
```

### Donation Amount Distributions: Treatment vs. Control (Donors Only)

The two histograms above show the distribution of donation amounts among individuals who made a donation, separated by treatment assignment. The red vertical line in each plot marks the **average donation amount** within that group.

- In the **treatment group**, the distribution of donations is right-skewed, with most donations falling between $10 and $100. The average donation is just below $44.
- In the **control group**, the pattern is similar, though the average donation is slightly higher, around $45.5.

Visually, the distributions are quite similar. Both groups show large clustering at common donation levels (e.g., $25, $50, $100), suggesting donors may be responding to standard ask amounts or psychological anchors.

### Interpretation

These plots support the earlier statistical analysis: **being offered a match does not appear to increase the average amount donated among those who give**. In fact, the average donation is marginally lower in the treatment group. However, this difference is not statistically significant, and the shapes of the two distributions are nearly identical.

This visualization reinforces the paper’s broader conclusion that matching gifts increase the **likelihood of giving**, but **not the size** of the donation once that decision has been made. The behavioral nudge seems to influence the extensive margin (give vs. not give), not the intensive margin (how much to give).

## Simulation Experiment

As a reminder of how the t-statistic "works," in this section I use simulation to demonstrate the Law of Large Numbers and the Central Limit Theorem.

Suppose the true distribution of respondents who do not get a charitable donation match is Bernoulli with probability p=0.018 that a donation is made. 

Further suppose that the true distribution of respondents who do get a charitable donation match of any size  is Bernoulli with probability p=0.022 that a donation is made.

### Law of Large Numbers

```{r}

# Set seed for reproducibility
set.seed(123)

# Simulate 100,000 control group donations (mean = 0.813, sd = 10)
control <- rnorm(100000, mean = 0.813, sd = 10)

# Simulate 10,000 treatment group donations (mean = 0.966, sd = 10)
treatment <- rnorm(10000, mean = 0.966, sd = 10)

# Calculate 10,000 differences (each treatment value - sampled control value)
diffs <- treatment - sample(control, size = 10000)

# Compute cumulative average
cum_avg <- cumsum(diffs) / seq_along(diffs)

# Create data for plotting
plot_data <- data.frame(
  x = 1:10000,
  cum_avg = cum_avg
)

# Plot the convergence
library(ggplot2)
ggplot(plot_data, aes(x = x, y = cum_avg)) +
  geom_line(linewidth = 0.8, color = "steelblue") +
  geom_hline(yintercept = mean(treatment) - mean(control), color = "red", linetype = "dashed") +
  coord_cartesian(ylim = c(-1, 1)) +  # focus the Y-axis for cleaner view
  labs(
    title = "Cumulative Average of Treatment-Control Differences",
    subtitle = "Simulating the Law of Large Numbers",
    x = "Sample Size",
    y = "Cumulative Average Difference"
  ) +
  theme_minimal()
```
### Simulating the Law of Large Numbers

The plot above shows the cumulative average of 10,000 differences between randomly drawn treatment and control values. Specifically, we simulate 100,000 values from the control distribution and 10,000 from the treatment distribution, then compute the difference in means one pair at a time. The blue line shows the evolving average of those differences as the sample size increases.

At the beginning of the plot, we see high variability — the cumulative average fluctuates wildly with the first few draws. This reflects the influence of **random noise** when the sample size is small. However, as the number of observations increases, the cumulative average begins to **stabilize**, eventually converging toward the red dashed line, which represents the **true difference in means** between the treatment and control groups.

This behavior is a textbook demonstration of the **Law of Large Numbers**: as sample size increases, the sample average becomes a more accurate estimate of the population average. In practical terms, it tells us that while individual comparisons may be noisy, large samples can reveal reliable patterns. Here, the cumulative average confirms the treatment group tends to donate more than the control group — and that this effect emerges clearly as sample size grows.

### Central Limit Theorem

```{r}

set.seed(123)

library(ggplot2)
library(dplyr)

# Simulate sampling distribution of differences
simulate_diff_means <- function(n, reps = 1000, mu_c = 0.813, mu_t = 0.966, sd = 10) {
  diffs <- replicate(reps, {
    mean(rnorm(n, mean = mu_t, sd = sd)) - mean(rnorm(n, mean = mu_c, sd = sd))
  })
  data.frame(diff = diffs, sample_size = paste0("n = ", n))
}

# Generate datasets
sizes <- c(50, 200, 500, 1000)
sim_data <- do.call(rbind, lapply(sizes, simulate_diff_means))

# Plot histograms
ggplot(sim_data, aes(x = diff)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  facet_wrap(~sample_size, scales = "free", ncol = 2) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Sampling Distribution of Differences in Means",
    subtitle = "Simulated at Varying Sample Sizes",
    x = "Difference in Means (Treatment - Control)",
    y = "Count"
  ) +
  theme_minimal()
```

### Sampling Distributions at Different Sample Sizes

The four histograms above show the **sampling distribution of the difference in means** between treatment and control groups at sample sizes of 50, 200, 500, and 1000. Each plot represents 1,000 simulations where, for each simulation, we draw `n` observations from each group, compute their sample means, and record the difference.

- At **n = 50**, the distribution is wide and noisy, with a substantial spread. In this case, **zero is near the center** of the distribution, meaning we would often observe a difference close to zero just by chance.
  
- At **n = 200**, the distribution is more concentrated, though still displays some variability. **Zero is still within the range of plausible outcomes**, but less frequently near the center.
  
- At **n = 500**, the sampling distribution narrows significantly, and **zero starts to move toward the edge** of the distribution — suggesting stronger evidence of a real difference.
  
- At **n = 1000**, the distribution is tight and centered around the true mean difference. **Zero is now well into the tails**, meaning it’s unlikely we would observe a difference this close to zero if the treatment and control distributions were truly the same.

### What We Learn

These plots visually demonstrate the **power of larger sample sizes**. As the sample size increases:
- The **sampling variability decreases**
- The **sampling distribution becomes more precise**
- We are better able to detect small true differences in population means

This is a direct illustration of the **Central Limit Theorem** and the **Law of Large Numbers** in action. At larger sample sizes, we can distinguish signal from noise more effectively — and zero becomes an unlikely outcome when a real treatment effect exists.





